---
title: "Time Series Analysis Presentation Midterm"
author: "Group No. 15 - Tejesh Pawar, Anju Anand"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
editor_options: 
  chunk_output_type: console
---

# Introduction

##

In this presentation, we will analyze a multivariate time series dataset and perform modeling and forecasting using R.

</br>

Datasets chosen: 


**Gross Domestic Product (GDP)** dataset and the **Consumer Price Index (CPI) Energy** dataset.

</br>

We have chosen these datasets to explore the link between economic growth and energy prices, providing fresh insights for decision-making.

## 

### <span style="font-size: 40px;">Loading the required libraries:</span>

</br>

```{r, echo = TRUE}
# Load required libraries
library(datasets)
library(forecast)
library(ggplot2)
```

</br>

### <span style="font-size: 40px;">Loading both the datasets:</span>

</br>

```{r, echo = TRUE}
# Load GDP dataset
gdp_data <- read.csv("GDP.csv")

# Load Consumer Price Index Energy dataset
cpi_data <- read.csv("CUUR0000SASLE.csv")

```


## Plotting the GDP data

</br>

```{r}
# Convert the DATE column to Date type
gdp_data$DATE <- as.Date(gdp_data$DATE)

# Plot the GDP time series data
g <- ggplot(gdp_data, aes(x=DATE, y=GDP)) +
  geom_line() + # Line plot for the time series
  labs(title="GDP Over Time", x="Date", y="GDP (in billions)") + # Adding labels
  theme_minimal() # Minimal theme for a nicer look

# Display the plot
print(g)
```


## Plotting the CPI data

</br>

```{r}
# Convert the DATE column to Date type
cpi_data$DATE <- as.Date(cpi_data$DATE)

# Plot the CPI Energy time series data
g <- ggplot(cpi_data, aes(x=DATE, y=CUUR0000SASLE)) +
  geom_line() + # Line plot for the time series
  labs(title="CPI Energy Over Time", x="Date", y="CPI Energy Index") + # Adding labels
  theme_minimal() # Minimal theme for a nicer look

# Display the plot
print(g)
```

## GDP Data - Empirical Data Analysis

```{r}
# Summary of GDP dataset
summary(gdp_data)

# converting to numeric values
gdp_values <- as.numeric(gdp_data$GDP)

# Plotting GDP data against dates
plot(as.Date(gdp_data$DATE), gdp_values, type = "l", xlab = "Date", ylab = "GDP", main = "GDP Time Series")
```

::: {style="font-size: 70%;"}


* This dataset uses a range of numbers to summarise GDP growth over a significant amount of time, most likely decades.

* The lowest and maximum GDP numbers indicate a definite upward tendency in the expansion of the economy.

* Since the median number is significantly greater than the minimum, most years had GDPs that were higher than the lowest.

* The difference between the maximum and median values indicates that certain years saw significantly more growth than the average.

:::

## GDP data - ACF, PACF Plots & Diagnostics

```{r}
# Plot ACF and PACF of GDP data
par(mfrow = c(1, 2))  
acf(gdp_values, main = "ACF of GDP Data")
pacf(gdp_values, main = "PACF of GDP Data")
```


## GDP data - ACF, PACF Plots & Diagnostics

</br>


::: {style="font-size: 72%;"}

* The ACF plot at lag 1 indicates a strong correlation between current and immediate past GDP, suggesting the potential influence of the latter on the former.

* The diminishing spikes in the ACF plot with increasing lags hint at possible seasonality, which weakens over time.

* A significant rise at lag 1 in the PACF plot supports using an AR(1) model for GDP forecasting, utilizing the previous value plus an error term.

* The absence of significant spikes at higher lags in the PACF suggests that more complex autoregressive models might be unnecessary.


:::



## CPI Energy Consumption - Empirical Data Analysis

```{r}
# Summary of CPI dataset
summary(cpi_data)

# Convert DATE column to Date class
cpi_data$DATE <- as.Date(cpi_data$DATE)

# Plot CPI data against dates
plot(cpi_data$DATE, cpi_data$CUUR0000SASLE, type = "l", xlab = "Date", ylab = "CPI", main = "CPI-Energy services Time Series")
```



::: {style="font-size: 70%;"}

* This large dataset examines the Consumer Price Index for Energy Services (CUUR0000SASLE) and spans probably more than 80 years.

* The huge gap between the lowest and greatest numbers (21.4 vs. 408.1) points to a long-term, substantial increase in the cost of energy services.

* A skewed distribution is shown by the median (142.8) being lower than the mean (155.8), with more years having lower CUUR0000SASLE values.


:::



## CPI-Energy Consumption ACF, PACF Plots & Diagnostics

```{r}
# Plot ACF and PACF of CPI data
par(mfrow=c(1,2))
acf(cpi_data$CUUR0000SASLE, main = "ACF for CPI Data")
pacf(cpi_data$CUUR0000SASLE, main = "PACF for CPI Data")
```


## CPI-Energy Consumption ACF, PACF Plots & Diagnostics

</br>

::: {style="font-size: 80%;"}

* An effective plot shows errors distributed evenly along a straight line, suggesting a normal, bell-shaped distribution.

* Clustering of points should be avoided, indicating unbiased error distribution.

* The absence of curves or peculiar patterns in the line points to normally distributed errors.

* While a good visual plot is encouraging, further analysis is necessary to confirm the errors' normal distribution.

:::



# Modeling

## 

</br>

To address the task given for GDP and Consumer Price Index (CPI) Energy datasets using time series modeling, we can consider using two commonly employed models for such data: 

* **Autoregressive Integrated Moving Average (ARIMA) model** 
</br>and</br>
* **State Space Model**


## ARIMA Model:
</br>

##### Notation:

- $Y_t$: the time series data at time $t$
- $\nabla Y_t$: the differenced series (e.g., $Y_t - Y_{t-1}$) to achieve stationarity
- $p$: the number of lag observations included in the model (autoregressive terms)
- $d$: the degree of differencing
- $q$: the size of the moving average window


## Relevant Mathematical Background:

</br>

- The model is defined by the equation: $\nabla^d Y_t = \alpha + \sum_{i=1}^{p} \phi_i \nabla^d Y_{t-i} + \sum_{i=1}^{q} \theta_i \epsilon_{t-i} + \epsilon_t$
- Where $\alpha$ is a constant, $\phi_i$ are the parameters for the autoregressive terms, $\theta_i$ are the parameters for the moving average terms, and $\epsilon_t$ is white noise.



## State Space Model
</br>

#### Notation:

- $Y_t$: the time series data at time $t$
- $\alpha, \beta, \gamma$: smoothing parameters for the level, trend, and seasonal components, respectively
- $l_t$: level at time $t$
- $b_t$: trend at time $t$
- $s_t$: seasonal component at time $t$

##  Relevant Mathematical Background:

</br>

- The model components are updated as follows:
  - Level: $l_t = \alpha (Y_t - s_{t-L}) + (1 - \alpha)(l_{t-1} + b_{t-1})$
  - Trend: $b_t = \beta (l_t - l_{t-1}) + (1 - \beta)b_{t-1}$
  - Seasonality: $s_t = \gamma (Y_t - l_{t-1} - b_{t-1}) + (1 - \gamma)s_{t-L}$
- Where $L$ is the length of the seasonal cycle.


## Application to Datasets:

</br>

::: {style="font-size: 75%;"}

For GDP and CPI Energy datasets, we can follow these steps:

:::

</br>


**<span style="font-size: 50px;">1. Simulation to Demonstrate Theory:</span>**


::: {style="font-size: 75%;"}

  * Generate synthetic time series data that resembles the GDP and CPI Energy datasets.
  * Apply the ARIMA and state space model to this synthetic data to demonstrate how the models capture various components like trends and seasonal effects.

:::
  
## 

</br>


**<span style="font-size: 50px;">2. Fit Model to Your Dataset:</span>**

</br>

::: {style="font-size: 75%;"}

  * Use historical data for GDP and CPI Energy to fit the ARIMA and state space models.
  * Determine the optimal parameters (e.g., *p,d,q* for ARIMA) using techniques like grid search and AIC/BIC minimization.
  - For the state space model, estimate the smoothing parameters and seasonal components.

:::
 
## 


**<span style="font-size: 50px;">3. Discuss Model Fit Diagnostics:</span>**


</br>

::: {style="font-size: 75%;"}

  - Analyze residuals to check for any patterns or non-randomness.
  - Use diagnostic plots like the autocorrelation function (ACF) and partial autocorrelation function (PACF) plots for ARIMA residuals.
  - For the state space model, check if the level, trend, and seasonal components accurately capture the corresponding elements in the data.
  - Compare the models using information criteria and out-of-sample prediction performance to assess their goodness of fit.
  
</br>
  
The above outlines a general approach to modeling GDP and CPI Energy datasets using time series methods. 

:::



## GDP Data Simulation and Arima Model fit

</br>

```{r, echo = TRUE}
# Quick simulation for ARIMA model
set.seed(123)
arima_sim <- arima.sim(model = list(order = c(1,1,1), 
                                    ar = 0.7, ma = -0.5), n = 100)

```



```{r}
# Convert GDP data to a time series object
gdp_ts <- ts(gdp_data$GDP, start = c(1947, 1), frequency = 4)

```

</br>

```{r, echo = TRUE}
# Fit ARIMA model to GDP time series
arima_model_gdp <- auto.arima(gdp_ts)
```


## <span style="font-size: 50px;"> Diagnostic statistics of ARIMA model </span>

```{r}
# Diagnostic statistics of ARIMA model
checkresiduals(arima_model_gdp)
```

## 

</br>

::: {style="font-size: 68%;"}

* A proper residuals plot with errors randomly centered around zero suggests no underlying patterns, indicating the model's accuracy without consistent bias.

* Residuals evenly spread around zero reflect the model's effectiveness in capturing the data's trend, avoiding systematic over or underestimation.

* The absence of outliers is crucial, as they can point to external factors unaccounted for by the model.

* The residuals plot serves as a tool for identifying improvement areas and assessing the ARIMA model's fit.

* The Ljung-Box test result implies the residuals are likely random, supporting model validity.

* A high p-value (0.40) indicates minimal concerns about residual randomness.

* Collectively, these insights hint at the ARIMA(0,2,2) model being a potentially suitable choice.

:::

## Simulation and State Space Model fit on CPI-Energy Consumption

</br>

```{r, echo = TRUE}
# Convert CPI column to time series object
cpi_ts <- ts(cpi_data$CUUR0000SASLE, start = min(cpi_data$DATE), frequency = 12)  

```

</br>

```{r, echo = TRUE}
# Fit state space model to energy consumption data
stsm_model_cpi <- StructTS(cpi_ts)

```

## <span style="font-size: 50px;"> Diagnostic statistics of State Space model</span>


```{r}
# Diagnostic statistics of State Space model
checkresiduals(stsm_model_cpi)
```

## 

</br>
</br>

::: {style="font-size: 80%;"}

* The Ljung-Box test indicates potential issues with a high Q* and low p-value.

* This suggests the model's residuals are not random but show dependency, indicating autocorrelation.

* The influence of past errors on current errors isn't being sufficiently addressed by the model.

* The autocorrelation concern persists across multiple time steps (24 lags).

* These findings suggest the need for adjustments to the Structural Time Series model to address the non-random error behavior.

:::



# Forecasting


## GDP Forecasting Using ARIMA

</br>

```{r}
# Forecasting using ARIMA model for GDP
arima_forecast <- forecast(arima_model_gdp, h = 12)

# Print forecasts
print("ARIMA Forecast for GDP:")
print(arima_forecast)

```


## Plot ARIMA forecast for GDP

</br>


```{r}
plot(arima_forecast, main = "ARIMA Forecast for GDP", xlab = "Year", ylab = "GDP Value")
legend("topleft", legend = "Forecast", col = "blue", lty = 1, inset = c(0, 0), xpd = TRUE)

```


## CPI-Energy Forecasting using State Space

</br>

```{r}
# Convert DATE column to Date format
cpi_data$DATE <- as.Date(cpi_data$DATE)

# Create a time series object with correct start and frequency
cpi_ts <- ts(cpi_data$CUUR0000SASLE, start = c(1957, 1), frequency = 12)

# Fit state space model
ss_model <- StructTS(cpi_ts, type = "BSM")

# Forecast using the state space model
forecast_values <- forecast(ss_model, h = 12)

# Print forecasts
print("State Space Model Forecast for CPI Energy:")
print(forecast_values)
```

## Plot State Space forecast for CPI Energy

</br>
```{r}
# Plot State Space Model forecast for CPI Energy
plot(forecast_values, main = "State Space Model Forecast for CPI Energy", xlab = "Year", ylab = "CPI Energy Index")
legend("topleft", legend = "Forecast", col = "blue", lty = 1, inset = c(0, 0), xpd = TRUE)

```
</br>

## Compare forecasts using cross-validation for multiple time horizons:

</br>

::: {style="font-size: 70%;"}

* The GDP forecast details three-year monthly economic growth; the ten-month CPI energy projection highlights energy consumption changes.

* CPI energy targets consumption trends; GDP provides broader economic insights.

* Comparing both offers insights into economic-energy dynamics over varying periods.

* Joint analysis of these forecasts presents a holistic economic and energy perspective.
:::

## Discuss the results and compare the two models:

</br>

::: {style="font-size: 70%;"}

* The three-year GDP forecast suggests robust economic growth, implying a positive outlook.

* Conversely, the ten-month CPI energy forecast predicts a decrease in energy consumption.

* The CPI energy model highlights shifts in energy demand, while the GDP model focuses on economic health.

* Analyzing both forecasts aids stakeholders in making informed decisions regarding energy and economic strategies.

* Together, the forecasts offer a multifaceted view of the economic and energy sectors.

:::

## Discuss practical implications of these forecasts in the context of the data:

</br>

::: {style="font-size: 70%;"}

* The GDP estimate helps firms and politicians make long-term investment and economic planning decisions.

* The CPI energy forecast helps energy firms and policymakers manage resources and develop energy policies.

* By offering a comprehensive view of economic and energy trends, combining the two predictions improves decision-making.

* Through early planning and efficient resource allocation, stakeholders can leverage the forecasts to foresee future possibilities and difficulties.

:::

# Conclusion

## 

</br>
</br>

::: {style="font-size: 70%;"}
* The GDP forecast guides long-term strategic planning for businesses and policymakers.

* The CPI energy forecast supports energy sector resource management and policy development.

* Merging the forecasts strengthens decisions with insights into economic and energy trends.

* These insights enable stakeholders to plan ahead and optimize resource use, preparing for future scenarios.

::: 

# Questions?


##
</br>
</br>
</br>
</br>
<span style="display:block; text-align:center; font-size: 110px;">**THANK YOU!**</span>

```{r}
```
